---
output: github_document

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
options("progress_enabled" = FALSE)
```

# cito

[![Project Status: Active -- The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active) [![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0) [![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/cito)](https://cran.r-project.org/package=cito) [![R-CMD-check](https://github.com/citoverse/cito/workflows/R-CMD-check/badge.svg)](https://github.com/citoverse/cito/actions)

<!-- badges: end -->

'cito' aims at helping you build and train Neural Networks with the standard R syntax. It allows the whole model creation process and training to be done with one line of code. Furthermore, all generic R methods such as print or plot can be used on the created object. It is based on the 'torch' machine learning framework which is available for R. Since it is native to R, no Python installation or any further API is needed for this package.

## Installation

Before installing 'cito' make sure 'torch' is installed. See the code chunk below if you are unsure on how to check this

```{r, eval = FALSE}
# check package 
if(!require('torch',quietly = TRUE)) install.packages('torch')
library('torch') 

#install torch
if(!torch_is_installed()) install_torch()
```

If you have trouble installing 'torch', please [visit their website](https://torch.mlverse.org/docs/articles/installation.html) or create an issue on [our github website](https://github.com/citoverse/cito/issues). We are happy to help.

A stable version of cito from CRAN can be installed with:

```{r, eval=FALSE}
install.packages("cito")

```

The development version from [GitHub](https://github.com/) is available with:

```{r, eval=FALSE}
if(!require('devtools', quietly = TRUE)) install.packages('devtools')
devtools::install_github('citoverse/cito')
```

## Example

Once installed, the main function `dnn()` can be used. See the example below. A more in depth explanation can be found in the vignettes.

1.  Fit model with bootstrapping (for uncertainties)

```{r}
library(cito)
nn.fit <- dnn(Sepal.Length~., data = datasets::iris, bootstrap = 30L)
```

2.  Check if models have converged (compare training loss against baseline loss (=intercept only model)):

```{r, eval=FALSE}
analyze_training(nn.fit)
# At 1st glance they are converged since their loss is lower than the baseline loss.
```

3.  Plot model architecture

```{r}
plot(nn.fit)
```

4.  Return xAI effects and their uncertainties:

```{r}
summary(nn.fit)
```

5.  Predict (with confidence intervals):

```{r}
dim(predict(nn.fit, newdata = datasets::iris))
```
