---
title: "Advanced: Custom loss functions and prediction intervals"
author: "Maximilian Pichler"
date: "`r Sys.Date()`"
abstract: "This vignette shows how cito can be used for advanced techniques such as custom loss functions."
output:
 rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    html_document:
      toc: true
      theme: cerulean
vignette: >
  %\VignetteIndexEntry{Advanced: Custom loss functions and prediction intervals}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi=32,
  out.width="400px",
  fig.align="center"
)
options("progress_enabled" = FALSE)

```

## Custom loss functions

We can pass custom loss functions to cito. R variables/values that are used within the loss function and that should be additionally optimized must be passed to cito via the custom_parameters argument in `dnn(...custom_parameters = list(name_of_parameter=...))`

Examples:

-   (Complex) likelihood functions
-   Advanced: Quantile regression

Requirements: - Complex calculations have to be written in torch - All functions/calls must have derivatives.

### Example 1: Custom (likelihood/loss) functions

Gaussian likelihood (already implemented, but still a nice example). Custom parameters must be passed as a list to the custom_parameters function. The names must match the names of the parameters in the custom loss function. The values of the named custom parameters will be the initial values. Cito will automatically convert them to torch tensors:

```{r,eval=TRUE}
library(cito)
library(torch)
gaussian_ll = function(pred, true, ...) {
  loss = -torch::distr_normal(pred, scale = torch::torch_exp(scale_par))$log_prob(true)
  return(loss$mean())
}

# Simulate some data
X = runif(200)
Y = 2*X + rnorm(200, sd = 0.4)
df = data.frame(X = X, Y = Y)

m = dnn(Y~X, data = df, 
        loss = gaussian_ll, # custom function 
        verbose = FALSE,
        custom_parameters = list(scale_par = 0.0) # custom parameter that should be addtionally optimized
        )
```

The optimized parameters are saved in the parameter field:

```{r}
exp(m$parameter$scale_par) # true scale parameter: 0.4!
```

### Example 2: Quantile regression

The bootstrapping approach provides confidence intervals, but not prediction intervals. We could use likelihoods, such as the Gaussian likelihood, to fit a constant prediction interval. However, we often use loss functions, such as the mean squared error in ML/DL, which don't have an intrinsic parametrization for prediction intervals. We can approximate prediction intervals with quantile regression (a form of conformal prediction), which has the advantage of providing constant prediction intervals, which can be beneficial in certain situations (e.g., heteroscedasticity):

Simulate data:

```{r}
sim_in = function(n = 5) {
  S = diag(1., 3)
  S[1,2]=S[2,1]=0.0
  X = mvtnorm::rmvnorm(n, sigma = S)
  X1 = X[,1]
  C = X[,2]
  X2 = X[,3]
  Y = 1*X1 + 0.1*X2 + 0.0*C + rnorm(n, sd = 0.3+2*1.8^(X1+1))
  return(data.frame(Y = Y, X1 = X1, X2 = X2, C = C))
}

data = sim_in(500L)
plot(data$X1, data$Y)
```

The variance increases with higher feature values

Quantile Regression:

```{r,eval=TRUE}
library(torch)

q1 = torch_tensor(0.05)
q2 = torch_tensor(0.5)
q3 = torch_tensor(0.95)
loss_func = function(pred, true,...) {
  l1 = torch_max(q1*(true[,1,drop=FALSE]-pred[,1,drop=FALSE]), other = (1.0-q1)*(pred[,1,drop=FALSE]-true[,1,drop=FALSE]))
  l2 = torch_max(q2*(true[,2,drop=FALSE]-pred[,2,drop=FALSE]), other = (1.0-q2)*(pred[,2,drop=FALSE]-true[,2,drop=FALSE]))
  l3 = torch_max(q3*(true[,3,drop=FALSE]-pred[,3,drop=FALSE]), other = (1.0-q3)*(pred[,3,drop=FALSE]-true[,3,drop=FALSE]))
  return(l1+l2+l3)
}


m = dnn(cbind(Y, Y, Y)~., data = data, 
        lr = 0.01,
        loss = loss_func,
        lambda = 0.000, alpha = 0.5,
        epochs = 70L, hidden = c(30L, 30L), 
        activation = "selu", verbose = TRUE, plot = FALSE)

plot(data$X1, data$Y)
lines(smooth.spline(data$X1, predict(m)[,1], spar = 0.01), col = "blue")
lines(smooth.spline(data$X1, predict(m)[,3], spar = 0.01), col = "blue")
lines(smooth.spline(data$X1, predict(m)[,2], spar = 0.01), col = "red")
```

### Example 3: Using cito for optimization

We can use neural networks in an unconventional way to optimize arbitrary functions - as long as we can compute the analytic derivative of the function using torch. We need to provide the function to be optimized as a series of Torch operations. Our model will predict the parameters (based on noise, the inputs don't matter), in the custom loss function we will then use the model function (we want to optimize) to compute the loss and return it to the optimizer. In that way we overfit to the noisy inputs and the DNN will learn to predict the optimal set of parameters - independent of the input.

```{r}
X = runif(200)
Y = 2*X + rnorm(200, sd = 0.4)
df = data.frame(X = X, Y = Y)

# Function we want to optimize (linear model)
Xt = torch_tensor(matrix(X))
Yt = torch_tensor(matrix(Y))

model_lm = function(par) {
  pred = Xt$matmul(par[,1,drop=FALSE])
  loss = -torch::distr_normal(pred, scale = torch::torch_exp(par[,2,drop=FALSE]))$log_prob(Yt)
  return(loss$mean())
}

custom_loss = function(pred, true, ...) {
  if(nrow(pred) > 1) return(torch_zeros(1L)) # disable loss calculation
  loss = model_lm(pred)
  return(loss)
}

# X and Y values don't matter, number of columns in Y has to match the number of parameters we want to optimize
noise = matrix(runif(300*5), 300, 5)
noise_y = matrix(runif(300*2), 300, 2)
df = data.frame(y1 = noise_y[,1], y2 = noise_y[,2], noise)

m = dnn(cbind(y1, y2)~., data = df, loss = custom_loss, batchsize = 1L, epochs = 20L)
# Effect:
mean(predict(m)[,1])
# SD
mean(exp(predict(m)[,2]))
```
